---
title: "A CSV of bicycle rental transactions from the Bay Area Bike Share Data Challenge."
author: "Toby Lam"
date: "Wednesday, January 14, 2015"
output: html_document
---

##Background

The dataset would be download from this [1]. Suppose the file was downloaded in the machine. Let's have a brief look about the data.

```{r}
#read the file
df = read.csv("bike_trip_data.csv", stringsAsFactor=FALSE)

#checking
dim(df) #dimension of the data
head(df)
summary(df)
```

The dataset contains 144015 rows and 11 columns. The columns are listed as follows:

- Trip.ID
- Duration (in seconds)
- Start.Date
- Start.Station
- Start.Terminal - a number which maps the Start.Station
- End.Date
- End.Station
- End.Terminal - a number which maps the End.Station
- Bike..
- Subscription.Type
- Zip.Code

Let's check the number of unique start-station and unique end-station. Furthermore, check if they  contain the same set of the stations.
```{r}
length(unique(df$Start.Station))
length(unique(df$End.Station))

#check if they are with the same set of the station
any(sort(unique(df$Start.Station)) != sort(unique(df$End.Station)))
```

There are total 69 bike stations. In addition, the start-stations and end-stations are using the same set of stations.

##Solution to 3 problems
There are 3 questions that we need to answer. I listed the questions as follows:

1. What was the average total time (in minutes) used by a bicycle in the data?
2. What was the most popular day by trip frequency in this dataset?
3. (harder): Assuming there are 30 bikes per station, find what date and time the bikes FIRST need to be rebalanced. As in, there are 0 bikes at a terminal for a customer to rent. 
3. (easier): Assuming there are 30 bikes per station, find what date the bikes FIRST need to be rebalanced. As in, there are 0 bikes at a terminal for a customer to rent. Do this ignoring "Start.Date" and "End.Date" columns.

### 1. What was the average total time (in minutes) used by a bicycle in the data?

To answer this question, we only need to get the mean of the column "Duration".

```{r}
avgDuration = mean(df$Duration) / 60 #unit in minute
```

The average total time (in minutes) used by a bicycle in the data: _`r avgDuration`_ minutes

### 2. What was the most popular day by trip frequency in this dataset?

Refer to this question, I consider the most popular day by trip frequency is the most number of people who rented a bicycle in one day.

For people who rented a bicycle with a duration more than 24 hours, I only considered the frequency as 1 for the start day.

```{r}
#create a new data frame with the day only
t = as.data.frame(as.Date(df$Start.Date, "%m/%d/%y"))
names(t)[1] = "day"
idx = which.max(table(t))
popularDay = table(t)[idx]
```


As you can see, the most popular day by trip frequency is _`r names(popularDay)`_ which have _`r popularDay`_ number of people who rented a bicycle.


### 3. Assuming there are 30 bikes per station, find what date and time the bikes FIRST need to be rebalanced. As in, there are 0 bikes at a terminal for a customer to rent. 

I tried to tackle the harder part of this question. 

I created a function `getRebalanceDate` to obtain the rebalance time for a particular station (id). IDEA of the function:

- get all data frame (df) and the station id (id) from the parameters
- extract the rows (with columns: Start.Date only) which a bike was borrowed from the corrsponding id and store to a data frame (outDF)
- add a new column named state which all the entries with value -1
- extract the rows (with columns: End.Date only) which a bike was returned to the corrsponding id and store to a data frame (inDF)
- add a new column named state which all the entries with value 1
- combine two data frames together and store in a new data frame named DF
- sort the DF with respect to the date time in ascending order
- reset the state in first row by using the formula: 30 + the current state value 
- calcuate the acculated sum and check the first location with state value as ZERO
- if so, return such date time; otherwise, return NA

```{r}
getRebalanceDate = function(df, id){
  #create data frame - out
  outDF = as.data.frame(df[which(df$Start.Terminal == id), c("Start.Date")])
  outDF$state = -1
  names(outDF)[1] = "time"
  #create data frame - in
  inDF = as.data.frame(df[which(df$End.Terminal == id), c("End.Date")])
  inDF$state = 1
  names(inDF)[1] = "time"
  
  #combine two data frame
  DF = rbind(outDF, inDF)
  #sort by date time
  DF = DF[order(as.POSIXlt(DF$time, format="%m/%d/%y %H:%M")), ]
  #at the begining with 30 bicycle
  DF[1, "state"] = DF[1, "state"] + 30
  #calculate the acculated sum
  csum = cumsum(DF$state)
  #check which index get the value 0
  idxs = which(csum == 0)
  #get the date time
  as.character(DF[idxs[1], "time"])
}
```

To obtain the FIRST date and time need to rebalance the bikes, we only need to use lapply to call the getRebalanceDate function. Then, store the result in a data frame and sorted it by the time accordingly. The first element will be the FIRST date and time that needs rebalance.

```{r}
#create a record data frame
record = as.data.frame(as.character(unique(df$Start.Station)))
record$id = as.character(unique(df$Start.Terminal)) #get the ids

#get the rebalance time
l = unlist(lapply(record$id, function(x){
        getRebalanceDate(df, x)
      }))
#update the record
record$time = l

#sort the data with respect to the time
record = record[order(as.POSIXlt(record$time, format="%m/%d/%y %H:%M")), ]
```

The station _`r record[1, 1]`_ first need to re-balanced on _`r record[1, 3]`_.

There are a number of station required to rebalance also. We can get this information by using the following code.

```{r}
stationNeedRebalance = record[!is.na(record$time), ]
stationNeedRebalance
```

The total number of station need rebalance: _`r nrow(stationNeedRebalance)`_.

*Note:*

When I publish this document to Rpub, I encountered the errors about SSH and readBin. I refer to [2-3] to solve these two problems.

## Reference
[1] - http://mandrillapp.com/track/click/30315607/s3-us-west-1.amazonaws.com?p=eyJzIjoiRjBqdTctUHl3LTlMY1FkZU9SVVl3ZDRoV29VIiwidiI6MSwicCI6IntcInVcIjozMDMxNTYwNyxcInZcIjoxLFwidXJsXCI6XCJodHRwczpcXFwvXFxcL3MzLXVzLXdlc3QtMS5hbWF6b25hd3MuY29tXFxcL2RhdGF5ZWFyXFxcL2Jpa2VfdHJpcF9kYXRhLmNzdlwiLFwiaWRcIjpcIjFlMTRlMDVmYzNmNjRiNzE4NTEzYzQ1ZmNhNzI0OTA4XCIsXCJ1cmxfaWRzXCI6W1wiMTVlYzMzNWM1NDRlMTM1ZDI0YjAwODE4ZjI5YTdkMmFkZjU2NWQ2MVwiXX0ifQ

[2] - http://stackoverflow.com/questions/22537180/error-while-publishing-in-r-pubs

[3] - https://support.rstudio.com/hc/communities/public/questions/202448223-Publish-to-rpubs-Error-in-readBin-conn-what-raw-n-contentLength-invalid-n-argument